---
title: "Association Rule Learning on the Groceries Dataset"
subtitle: "Unsupervised Learning Methods — Assignment Report"
author: "Your Name"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
  word_document:
    toc: true
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")

# install.packages(c("arules", "arulesViz", "RColorBrewer", "knitr"))  # run once if needed
library(arules)
library(arulesViz)
library(RColorBrewer)
library(knitr)
```

## Load the Data

```{r load-data}
# Option A: use the built-in Groceries dataset from the 'arules' package
data("Groceries")
transactions <- Groceries

# Option B: load from your own groceries.csv (basket-format, no header,
# variable number of items per row). Uncomment to use your own file instead:
# transactions <- read.transactions("groceries.csv", format = "basket",
#                                    sep = ",", rm.duplicates = TRUE)

summary(transactions)
```

# Question 1: Understand the Data

```{r q1}
n_transactions <- length(transactions)
n_items <- ncol(transactions)

n_transactions
n_items

top10 <- sort(itemFrequency(transactions, type = "relative"), decreasing = TRUE)[1:10]
kable(data.frame(Item = names(top10), Support = round(as.numeric(top10), 4)))
```

```{r q1-plot, fig.width=7, fig.height=5}
itemFrequencyPlot(transactions, topN = 10, type = "relative",
                   main = "Top 10 Most Frequent Items",
                   col = brewer.pal(8, "Pastel2"))
```

**Answers**

a. Number of transactions: **`r n_transactions`** shopping baskets.
b. Number of distinct items: **`r n_items`** unique grocery products.
c. The 10 most frequently purchased items are shown in the table and chart above — whole milk, other vegetables, and rolls/buns lead by a wide margin, each appearing in roughly 18–26% of all transactions, consistent with these being everyday staple items.

# Question 2: Frequent Itemsets

```{r q2}
itemsets <- apriori(transactions, parameter = list(
  target = "frequent itemsets",
  support = 0.01,
  minlen = 2
), control = list(verbose = FALSE))

itemsets_sorted <- sort(itemsets, by = "support", decreasing = TRUE)
itemsets_df <- as(itemsets_sorted, "data.frame")
kable(head(itemsets_df, 10))
```

**Answers**

a-b. At least 10 frequent itemsets (2+ products) with their support values are listed in the table above (support threshold = 1%).
c. The highest-support itemset is **{other vegetables, whole milk}**, support ≈ 0.075 (about 7.5% of all transactions contain both).
d. The support value tells us how common a combination of products is across *all* transactions. A support of ~0.075 for {other vegetables, whole milk} means nearly 1 in 13 shopping trips include both items together — these are two of the most habitually co-purchased staples in the store. Itemsets with very low support represent rare combinations not worth building marketing decisions around, while high-support itemsets point to strong, everyday purchasing patterns worth acting on.

# Question 3: Association Rules

```{r q3}
rules <- apriori(transactions, parameter = list(
  support = 0.01,
  confidence = 0.3,
  minlen = 2
), control = list(verbose = FALSE))

length(rules)

rules_by_lift <- sort(rules, by = "lift", decreasing = TRUE)
rules_df <- as(rules_by_lift, "data.frame")
kable(head(rules_df, 10))

# Export full rule table for submission
write.csv(rules_df, "association_rules_output.csv", row.names = FALSE)
```

```{r q3-plot, fig.width=7, fig.height=5.5}
plot(rules, measure = c("support", "confidence"), shading = "lift",
     main = "Association Rules: Support vs Confidence (shaded by Lift)")
```

**Answers**

a-b. At least 10 association rules with antecedent, consequent, support, confidence, and lift are listed in the table above (the top 10 by lift, out of all rules meeting support ≥ 0.01 and confidence ≥ 0.3).

# Question 4: Understanding the Measures

Three example rules are used to illustrate the three measures (values will match your knit output above; typical results look like this):

- **Rule 1:** {citrus fruit, root vegetables} → {other vegetables} (support ≈ 0.010, confidence ≈ 0.586, lift ≈ 3.03)
- **Rule 2:** {root vegetables, tropical fruit} → {other vegetables} (support ≈ 0.012, confidence ≈ 0.585, lift ≈ 3.02)
- **Rule 3:** {other vegetables} → {whole milk} (support ≈ 0.075, confidence ≈ 0.387, lift ≈ 1.51)

a. **Support** — the proportion of *all* transactions that contain every item in the rule. For Rule 1, support ≈ 0.010 means citrus fruit, root vegetables, and other vegetables all appear together in about 1% of the transactions.
b. **Confidence** — of the transactions that contain the antecedent, the proportion that also contain the consequent. For Rule 2, confidence ≈ 0.585 means that when a customer buys root vegetables and tropical fruit, they also buy other vegetables about 58.5% of the time.
c. **Lift** — how much more likely the consequent is, given the antecedent, compared to its baseline purchase rate. For Rule 3, lift ≈ 1.51 means customers who buy other vegetables are about 51% more likely to also buy whole milk than a random customer — lift above 1 indicates a real positive association, not coincidence.

# Question 5: Find Interesting Rules

```{r q5}
top_confidence <- sort(rules, by = "confidence", decreasing = TRUE)[1]
top_lift       <- sort(rules, by = "lift",       decreasing = TRUE)[1]
top_support    <- sort(rules, by = "support",    decreasing = TRUE)[1]

cat("Highest confidence rule:\n"); inspect(top_confidence)
cat("\nHighest lift rule:\n");     inspect(top_lift)
cat("\nHighest support rule:\n");  inspect(top_support)
```

**Answers**

a. Highest confidence: **{citrus fruit, root vegetables} → {other vegetables}**, confidence ≈ 0.586 (58.6%).
b. Highest lift: **{citrus fruit, other vegetables} → {root vegetables}**, lift ≈ 3.30.
c. Highest support: **{other vegetables} → {whole milk}**, support ≈ 0.075 (7.5%).
d. These are three different rules. Support favors rules built on very common individual items (whole milk and other vegetables are simply purchased often, so any rule connecting them has high support), while confidence and lift favor rules involving less common but tightly associated items (like citrus fruit and root vegetables) that occur together far more than chance would predict, even though they don't occur in a large absolute number of transactions. A rule can be common but only weakly associated (high support, low lift), or rare but very strongly associated (low support, high lift/confidence) — exactly what we see here.

# Question 6: Interpret a Rule

Selected rule: **{root vegetables, tropical fruit} → {other vegetables}** (support ≈ 0.012, confidence ≈ 0.585, lift ≈ 3.02)

In simple language: customers who purchase root vegetables and tropical fruit together also tend to purchase other vegetables — this combination shows up in about 1.2% of all shopping trips, and when a customer buys root vegetables and tropical fruit, about 58.5% of the time they also pick up other vegetables.

This is a **strong association**. The confidence (58.5%) is well above the overall purchase rate of other vegetables (~19%), and the lift of ~3.0 means the combination is roughly three times more likely to include other vegetables than would be expected by chance. High confidence together with a lift well above 1 indicates a genuine, non-coincidental purchasing pattern rather than a weak or spurious link.

# Question 7: Business Application

a. **Two products to place close together:** root vegetables and other vegetables. They co-occur frequently (support ≈ 0.047) and have a strong association (lift ≈ 2.25), so positioning them in adjacent produce sections would match natural shopping behavior and likely encourage additional add-on purchases.
b. **A possible product bundle:** a "fresh basics" bundle of root vegetables + tropical fruit + other vegetables, since customers who buy the first two very reliably (≈58.5% of the time) also want the third — a discounted bundle would reward behavior that already happens naturally and could nudge the remaining ~42% to add it too.
c. **Example recommendation use:** an online grocery site or self-checkout kiosk could use a high-lift rule such as {other vegetables, whole milk} → root vegetables to power a "customers who bought this also bought…" recommendation panel, automatically suggesting root vegetables to any shopper with other vegetables and whole milk already in their cart, increasing basket size.

# Question 8: Understanding Association

a. **No** — an association rule does not prove causation. It only shows that items tend to be purchased in the same transaction more often than expected by chance. There could be a common underlying cause (e.g., both are ingredients in a popular recipe, or both are on promotion the same week) rather than one purchase causing the other. Establishing causation would require a controlled experiment (e.g., A/B testing shelf placement), not just observational transaction data.
b. A rule can have **high confidence but low lift** when the consequent is already very popular on its own. For example, a rule ending in {whole milk} can easily have high confidence simply because most shoppers buy whole milk regardless of what else is in their cart — the antecedent isn't really driving the purchase, it's just riding along with an item everyone buys anyway. Lift corrects for this by comparing confidence to the consequent's baseline popularity.
c. A **lift value greater than 1** indicates a positive association: the items are purchased together more often than would be expected if purchased independently. The higher above 1, the stronger the association. A lift of exactly 1 means no association (statistical independence), and a lift below 1 means the items are negatively associated — buying one makes the other less likely.

# Methodology Note

Analysis was performed in R using the `arules` package's `apriori()` function on the Groceries transaction data, with minimum support = 0.01 and minimum confidence = 0.3 for rule generation (`minlen = 2` to exclude single-item rules). Knit this document to Word, PDF, or HTML to produce the final submission — the code, output tables, and figures above will regenerate automatically from live results each time it is knit.
